---
title: "Congenital"
author: "Brian M. Schilder"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=TRUE, message=FALSE}
library(ggplot2)
```


[Are congenital phenotypes more enriched for fetal cell types?](https://github.com/neurogenomics/rare_disease_celltyping/issues/47)

- Are congenital phenotypes are more strongly enriched for fetal cell types than congenital phenotypes?
  - **Strategy**: Identify cell types that only occur in fetal samples (and not adult samples)
- Are congenital phenotypes are more strongly enriched for fetal cell types vs. adult cell types?
  - **Strategy**: Identify cell types that occur in both fetal and adult samples, and compare their relative enrichment levels in congenital vs. non-congenital disease.

# Import data

## Enrichment results
```{r}
res <- MultiEWCE::load_example_results("rare_disease_min_genes4_HumanCellLandscape.rds")
# split the Celltype columns via the first "_" only
res <- cbind(res,
      data.table::data.table(
        stringr::str_split(res$CellType,"_",n = 2, simplify = TRUE)) |> 
        `colnames<-`(c("stage","celltype")
                     )
      )
```

## Phenotype annotations
Use "congenital" column from GPT annotation.

```{r}
gpt <- HPOExplorer::gpt_annot_codify()
res_gpt <- data.table::merge.data.table(res, 
                                        gpt$annot,
                                        by="hpo_id")
```

# Analysis 

## Compare by stage

Each sample in the HCL comes from an "Adult", "Fetus", or "HESC" 
(human embryonic stem cell) stage. Here, we identify cell types that
are only present in fetal samples (which includes both Fetus or HESC)
but not adult samples.
We also identify cell types that are only present in adult samples.

Of the 63 unique cell types in HCL:
- Only 2 cell types are found only in fetal samples.
- Only 5 cell types are found only in adult samples.

```{r}
res_gpt[,list(celltype=unique(celltype)),keyby=stage]
fetal_stages <- c("Fetus","hESC")
fetal_only <- setdiff(res_gpt[stage %in% fetal_stages]$celltype,
                           res_gpt[stage=="Adult"]$celltype)
adult_only <- setdiff(res_gpt[stage=="Adult"]$celltype,
                      res_gpt[stage %in% fetal_stages]$celltype)
res_gpt[,stage_group:=ifelse(celltype %in% fetal_only,
                                "fetal_only",
                                ifelse(celltype %in% adult_only,
                                       "adult_only",
                                       "both")
                                )
        ]

cat("fetal_only:\n")
fetal_only
cat("adult_only:\n")
adult_only
```


Confusingly, these the proportion of "adult only" cell types are
more frequently associated with phenotypes that are "often" congenital 
relative to phenotypes that are "rarely" or "never congenital.

```{r}
ggplot(res_gpt[q<0.05,],
       aes(x=congenital_onset,
           fill=stage_group)) +
  geom_bar(position="fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y="% of significant results (FDR<=5%)") +
  theme_bw()  


ggplot(res_gpt[q<0.05 & stage_group!="both",],
       aes(x=congenital_onset,
           fill=stage_group)) +
  geom_bar(position="fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y="% of significant results (FDR<=5%)",
       title="'both' removed") +
  theme_bw()  

```


 

## Compare by substrings


Strangely, some cell types with the substrings "fetal"/"primordial"/"hESC" in them are found in the 
"Adult" samples as well. It's unclear whether this is:
- a mistake in the annotation
- a mistake during CTD processing
- a true biological phenomenon (e.g. circulating fetal cells in adult blood)

*NOTE*: I also tried using the substring "blast" but this only returned 1 cell type
("fibroblast") which is not a fetal cell type.

```{r}
fetal_keywords <- c("fetal","primordial","hESC")
res_gpt[,fetal_celltype:=grepl(paste(fetal_keywords,collapse="|"),
                               celltype,
                               ignore.case = TRUE)]
res_gpt[stage=="Adult" & fetal_celltype==TRUE,]$celltype |>
  unique()
```

If we compare only by celltype subtrings (regardless of sample stage), 
we find the expected relationship of a decreasing proportion of these "fetal" 
cell types with decreasing probability of the phenotype being congenital.

```{r}
ggplot(res_gpt[q<0.05,],
       aes(x=congenital_onset,
           fill=fetal_celltype)) +
  geom_bar(position="fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y="% of significant results (FDR<=5%)") +
  theme_bw()   

```



# Session Info

<details>
```{r}
sessionInfo()
```
</details>
