---
title: "Rare Disease Celltyping: reproducible manuscript"
format: html
editor: visual
---

```{r setup}
pkgs <- c("HPOExplorer","MSTExplorer")
for(x in pkgs){
  if(!require(x,character.only = TRUE)){
    BiocManager::install(paste0("neurogenomics/",x), dependencies=TRUE)
  }  
}
```

# Main analyses

## Phenotype-cell type associations

### Create gene x phenotype evidence matrix

```{r hpo_to_matrix, eval=FALSE}
## Create phenotype-evidence score matrix 
xmat <- HPOExplorer::hpo_to_matrix(formula = "gene_symbol ~ hpo_id")
```

### Run tests

```{r run_phenomix}
## Run phenomix on DescartesHuman CTD
lm_res1 <- MSTExplorer::run_phenomix(ctd_name = "DescartesHuman",
                                     annotLevel = 2,
                                     xmat = xmat)
## Run phenomix on HumanCellLandscape CTD
lm_res2 <- MSTExplorer::run_phenomix(ctd_name = "HumanCellLandscape",
                                     annotLevel = 3,
                                     xmat = xmat)
```

### Merge results

```{r Merge results}
results <- data.table::rbindlist(list(DescartesHuman=lm_res1,
                                      HumanCellLandscape=lm_res2),
                                idcol = "ctd")
## Apply multiple testing correction
results[,q:=stats::p.adjust(p,method="fdr")]
## Remove extra columns
results <- results[,-c("Effect","DFn","p<.05")]
remove(lm_res1,lm_res2)
data.table::setnames(results,
                     c("xvar","yvar","F"),
                     c("hpo_id","CellType","fold_change"))
```

### Upload results

Upload to `MSTExplorer` Releases on GitHub.

```{r Upload results, eval=FALSE}
f <- here::here("results/phenomix_results.tsv.gz")
data.table::fwrite(results,f)
piggyback::pb_upload(file = f, repo = "neurogenomics/MSTExplorer")
```

### Summarise results

Generate high-level summary statistics for the results, stratified by single-cell reference dataset.

```{r Summarise results}
results_summary <- MSTExplorer::summarise_results(results = results, 
                                                  save_path = here::here("tables/table1.csv"))
results_summary$tmerged
```

### Annotate results

```{r Annotate results}
## Add metadata 
results <- HPOExplorer::add_hpo_name(results)
results <- MSTExplorer::map_celltype(results)
```

## Symptom-cell type associations

```{r add_symptom_results, eval=FALSE}
## Run this after other steps (here for reference only)
# results <- MSTExplorer::add_symptom_results(results = results)
```

# Validation of expected phenotype-cell type relationships

## Summary of significant associations between phenotypes and cell types, aggregated by HPO branch

```{r plot_bar_dendro, fig.height=20}
out <- MSTExplorer::plot_bar_dendro(results = results, 
                                    save_path = here::here("figures/figure1.pdf"))
```

### More specific phenotypes are associated with fewer, more specific genes and cell types

```{r plot_ontology_levels}
ont_plots <- MSTExplorer::plot_ontology_levels(results = results, 
                                               save_path = here::here("figures/figure2.pdf"))
ont_plots$plot
```

### Hepatoblasts have a unique role in recurrent meningococcal disease.

```{r plot_bar_branches}
results_tmp <- HPOExplorer::add_ancestor(data.table::copy(results),
                                         lvl = 6,
                                         force_new = TRUE)
target_branches <- list("Recurrent bacterial infections"="leukocyte")
infections_out <- MSTExplorer::plot_bar_branches(
  results=results_tmp,
  target_branches=target_branches,
  facets = "hpo_name",
  legend.position="right",
  lvl=7,
  ncol=2,
  vbars="hepatoblast",
  facets_n=NULL,
  q_threshold=0.05,
  background_full=FALSE,
  save_path = here::here("figures/figure3.pdf"))
remove(results_tmp)
```

## Monarch Knowledge Graph recall

```{r}

```

# Annotation of phenotypes using generative large language models

## Import GPT annotations

```{r gpt_annot_codify}
gpt_annot <- HPOExplorer::gpt_annot_codify()
head(gpt_annot$annot)
```

## Add GPT annotations

```{r add_gpt_annotations}
results <- HPOExplorer::add_gpt_annotations(results, 
                                            annot = gpt_annot$annot)
```

## Congenital phenotypes are more often associated with fetal cell types

### Search for fetal substrings

```{r}
fetal_keywords <- c("fetal","fetus","primordial","hESC","embryonic")
results[,fetal_celltype:=grepl(paste(fetal_keywords,collapse="|"),
                               author_celltype,
                               ignore.case = TRUE)]
```

### Plot congenital vs. fetal/non-fetal

```{r ggbarstats}
ggbar <- ggstatsplot::ggbarstats(results[q<0.05 &  
                                         congenital_onset!="varies",],
                                  x="fetal_celltype",
                                  y="congenital_onset",
                                  package="palettetown",
                                  palette="mewtwo",
                                  proportion.test	= TRUE) 
ggplot2::ggsave(filename = here::here("figures/figure4.pdf"))
ggbar
```

# Diagnosis: Observed phenotypes/genotypes can be used to identify cell types in individuals

```{r predict_celltypes}
## Define input phenotypes/genes
phenotypes <- c("Scrotal hypospadias","Increased circulating progesterone")
genes_include <- c("MAGEL2","HERC2")
genes_exclude <- c("SNORD115-1")
## Predict cell types
ct <- MSTExplorer::predict_celltypes(phenotypes = phenotypes,
                                     genes_include = genes_include, 
                                     save_path = here::here("figures/figure5.pdf"))
```

# Prognosis: Cell types predict the probability of deadly diseases

## Add disease annotations

```{r add_death}
results <- HPOExplorer::add_death(results,
                                  allow.cartesian = TRUE,
                                  agg_by = c("disease_id","hpo_id"))
```

## Plot differential outcomes

```{r plot_differential_outcomes_heatmap}
## Count number of diseases associated with these phenotypes
phenotypes <- c("Hypotonia",
                "Neonatal hypotonia",
                "Generalized neonatal hypotonia")
phenotypes_ids <- HPOExplorer::map_phenotypes(phenotypes, to="id")
annot <- HPOExplorer::load_phenotype_to_genes(3)

diseases <- annot[hpo_id %in% phenotypes_ids]$disease_id|>unique()
message(length(diseases)," unique diseases are associated with: ",paste(phenotypes,collapse = "; "))

## Generate plot
differential_outcomes <- MSTExplorer::plot_differential_outcomes_heatmap(
  phenotypes = phenotypes,
  results = results,
  save_path = here::here("figures/figure6.pdf"),
  height=7, 
  width=15
)
differential_outcomes$plot
```

# Therapeutics: Gene therapy target identification

## Prioritise therapeutic candidates

```{r prioritise_targets}
targets <- MSTExplorer::prioritise_targets(results = results) 
```

### Prioritised target filtering steps

```{r plot_report}
report_plot <- MSTExplorer::plot_report(rep_dt = targets$report, 
                                        results = results, 
                                        save_plot = here::here("figures/figure7.pdf")) 
```

# Networks of prioritised cell type-specific gene targets

## Recurrent meningococcal disease

```{r prioritise_targets_network1}
vn1 <- MSTExplorer::prioritise_targets_network(
  top_targets = targets$top_targets[hpo_name=="Recurrent meningococcal disease"])
vn1
```

## Coma

```{r prioritise_targets_network2}
vn2 <- MSTExplorer::prioritise_targets_network(
  top_targets = targets$top_targets[hpo_name=="Coma"])
vn2
```

## Neonatal hypotonia

```{r prioritise_targets_network3}
vn3 <- MSTExplorer::prioritise_targets_network(
  top_targets = targets$top_targets[hpo_name=="Neonatal hypotonia"])
vn3
```

## Alzheimer's Disease

```{r prioritise_targets_network4}
vn4 <- MSTExplorer::prioritise_targets_network(
  top_targets = targets$top_targets[grepl("Alzheimer",hpo_name, ignore.case = TRUE)|
                                    grepl("Alzheimer",disease_name, ignore.case = TRUE)])
vn4
```

## Parkinson's Disease

```{r prioritise_targets_network5}
vn5 <- MSTExplorer::prioritise_targets_network(
  top_targets = targets$top_targets[grepl("Parkinson",hpo_name, ignore.case = TRUE)|
                                    grepl("Parkinson",disease_name, ignore.case = TRUE)])
vn5
```

# Validate therapeutic targets

## Therapeutics Target Database recall 

```{r ttd_check}
## Gene therapy only
ttd_targets <- MSTExplorer::ttd_check(top_targets=targets$top_targets, 
                                      drug_types = "Gene therapy",
                                      allow.cartesian = TRUE,
                                      save_path = here::here("figures/figure9.pdf"))
## All therapy types
ttd_all_targets <- MSTExplorer::ttd_check(top_targets=targets$top_targets,  
                                          allow.cartesian = TRUE,
                                          save_path = here::here("figures/figureS3.pdf"))

```

# Session info

\<details\>

```{r sessionInfo}
utils::sessionInfo()
```

\</details\>
