---
title: "Validation"
author: "Brian M. Schilder"
date: "`r Sys.Date()`"
output: html_document
---

**Tools**
- [scOntoMatch](https://github.com/Papatheodorou-Group/scOntoMatch/)
- [metacells](https://github.com/tanaylab/metacells)
- [SATURN](https://github.com/snap-stanford/SATURN)

```{r setup}
library(MSTExplorer)
library(ggplot2)
```

# Import data
## MSTExplorer results
```{r}
res <- MSTExplorer::load_example_results()  
## Map CellOntology IDs onto original names in results files
res <- MSTExplorer::map_celltype(res)
### Count number of unique celltypes per CTD/stage ###
# res[,list(N=length(unique(CellType))), by=c("ctd","stage")]

ctd_names <- unique(res$ctd)
# res_old <- MSTExplorer::load_example_results("Descartes_All_Results_extras.rds")
```

## scRNAseq

- [HumanCellLandscape](https://cellxgene.cziscience.com/collections/38833785-fac5-48fd-944a-0f62a4c23ed1)
- [DescartesHuman](https://cellxgene.cziscience.com/collections/c114c20f-1ef4-49a5-9c2e-d965787fb90c)
  - [1 million-cell subset](https://cellxgene.cziscience.com/collections/c114c20f-1ef4-49a5-9c2e-d965787fb90c)
```{r}
obj_list <- list()
obj_list[["DescartesHuman"]] <- readRDS("~/Downloads/db4d63ab-0ca3-4fa5-b0cf-34955227912d.rds")
obj_list[["HumanCellLandscape"]] <- readRDS("~/Downloads/1b7484e3-83a0-47fe-847e-54d811a2adae.rds")
```

## CellTypeDatasets
```R
ctd <- MSTExplorer::load_example_ctd(c("ctd_DescartesHuman.rds",
                                     "ctd_HumanCellLandscape.rds")) 
Xctd <- lapply(ctd_names, function(x){
  lvl <- res[ctd==x]$annotLevel[1]
  ctd[[x]][[lvl]]$specificity
}) |> do.call(what = Seurat::RowMergeSparseMatrices)
```

## Harmonise celltypes

Only 13 celltype IDs are shared between the two datasets, partly due to differing ontology level annotations. 

```{r}
report_matches <- function(obj_list,
                           col="cell_type"){
  inter <- lapply(names(obj_list), function(nm){
    x <- obj_list[[nm]]
    ct <- unique(x@meta.data[[col]])
    message(nm,":: ",col,": ",length(ct))
    return(ct)
  }) |> Reduce(f=intersect)
  message("Intersection: ",length(inter))
  inter
}
report_matches(obj_list, col="cell_type")
```

```{r})
lapply(obj_list, function(x){
  ct <- unique(x$cell_type)
  message("cell_type: ",length(ct))
  return(ct)
}) |> Reduce(f=intersect)
```

Thus, we must traverse the ontology to find the last common ancestor of each cell type.
We use `scOntoMatch` to do this.

```{r}
ont <- KGExplorer::get_ontology("cl", remove_rings = TRUE)
# obj_list_minimal = scOntoMatch::ontoMultiMinimal(obj_list = obj_list, 
#                                                  ont = ont,
#                                                  anno_col = "cell_type",
#                                                  onto_id_col = "cell_type_ontology_term_id")
# report_matches(obj_list, col="cell_ontology_base")

obj_list$DescartesHuman$CellType <- EWCE::fix_celltype_names(obj_list$DescartesHuman$Main_cluster_name, make_unique = FALSE)
obj_list$HumanCellLandscape$CellType <- EWCE::fix_celltype_names(obj_list$HumanCellLandscape$author_cell_type,
                                                                 make_unique = FALSE)

obj_list_matched <- scOntoMatch::ontoMultiMatch(
  obj_list = obj_list,
  ont = ont,
  anno_col = 'cell_type',
  onto_id_col = "cell_type_ontology_term_id")
report_matches(obj_list_matched, col="cell_ontology_mapped")

map <- lapply(obj_list_matched, function(x){
  x@meta.data |>
    dplyr::select(CellType, 
                  # cell_type,
                  # cell_type_ontology_term_id,
                  cell_ontology_mapped) |>
    unique()
}) |>
  data.table::rbindlist(idcol = "ctd", fill = TRUE) 
RES_MAPPED <- data.table::merge.data.table(
  RES,
  map,
  by= c("ctd","CellType"),
  allow.cartesian = TRUE
) 

```



# Compare results

## Across datasets: DescartesHuman vs HumanCellLandscape

Within the fetal data, is there a correlation between phenotype-celltype enrichment statistics (p-value, q-value, fold-change) across CTD references?
```{r}
# res <- lm_res
res2 <- res|> 
  # subset(stage=="Fetus") |>
  # subset(q<0.05)|>
  data.table::dcast.data.table(
  formula = hpo_id+stage+cell_ontology_mapped ~ ctd,
  fun.aggregate = mean,
  drop = TRUE,
  value.var = c("p","q","fold_change"))
RES2 <- RES2[complete.cases(RES2)][,test_id:=.I]
### All results
message(length(unique(RES2$cell_ontology_mapped))," comparable celltypes.")
message(length(unique(RES2$hpo_id))," comparable phenotypes.")

### Significant results in both CTDs
RES_SIG <- RES2[q_HumanCellLandscape<.05 & q_DescartesHuman<.05]
message(length(unique(RES_SIG$cell_ontology_mapped))," comparable celltypes (FDR<0.05).")
message(length(unique(RES_SIG$hpo_id))," comparable phenotypes (FDR<0.05).")
```

### All results (q<1)

This plot makes no distinction between stages in DescartesHumna (fetal only)
and HumanCellLandscape (Human Embyronic Stem Cells (HESC), Fetal, and Adult).
```{r}
RES2 |>
  ggstatsplot::ggscatterstats(x="p_HumanCellLandscape", 
                              y="p_DescartesHuman",
                              point.args = list(alpha=.01)) +
    ggplot2::geom_density_2d_filled(alpha=.85) 

```
### Signficant results (q<0.05)

```{r}
plot_cor <- function(dt,
                     x="fold_change_HumanCellLandscape",
                     y="fold_change_DescartesHuman",
                     trans='log10'){
gg <- dt |>
  # subset(stage=="Fetus")|>
  # subset(q_DescartesHuman<0.05 & q_HumanCellLandscape<0.05) |>
  ggplot(aes(x=!!sym(x),
                 y=!!sym(y)
  )) +
  geom_density_2d_filled() +
  geom_point(alpha=.2, color="white")+
  # geom_hex() +
  geom_smooth(method="lm") +
  ggpubr::stat_cor(color="white")+
  ggside::geom_xsidehistogram(fill="blue") +
  ggside::geom_ysidehistogram(fill="slateblue") +
  theme_bw()
  if(!is.null(trans)){
    gg <- gg + 
      scale_x_continuous(trans=trans) +
      scale_y_continuous(trans=trans) +
      labs(x=paste0(trans,"(",x,")"),
           y=paste0(trans,"(",y,")"))
  }
  return(gg)
}
```


```{r}
plot_cor(RES_SIG, 
         x="fold_change_HumanCellLandscape",
         y="fold_change_DescartesHuman")
```

### Precision-Recall

Compute the % of significant (FDR<5%) phenotype-celltype enrichment results recapitulated using each reference.
```{r}
compute_pr <- function(dt, 
                       ref_qcut=.05){
  ground_truth <- dt[q_DescartesHuman<ref_qcut]$test_id
  lapply(c(0.001,0.01,seq(0,1,.05)), function(pcut){
    r2 <- dt[q_HumanCellLandscape<pcut]
    TP <- r2$test_id %in% ground_truth |> length()
    TN <- !(r2$test_id %in% ground_truth) |> length()
    FN <- ground_truth[!(ground_truth %in% r2$test_id)] |> length()
    FP <- r2$test_id[!(r2$test_id %in% ground_truth)] |> length()
    data.table::data.table(
      # TP/TP+FP
      precision=TP/(TP+FP),
      # TP/TP+FN
      recall=TP/(TP+FN),
      # TP+TN/total
      accuracy=(TP+TN)/(TP+TN+FP+FN),
      ## Add pvalue threshold
      pvalue_threshold=pcut,
      ref_qcut=ref_qcut
    )
}) |> data.table::rbindlist()
}
```


```{r}
pr <- lapply(c(0.01,0.05,seq(0,1,.1)), function(ref_qcut){
  compute_pr(RES2, ref_qcut=ref_qcut)
})|> data.table::rbindlist()

pr |>
  ggplot(aes(x=recall, y=precision, 
             color=as.factor(pvalue_threshold))) +
  geom_line() +
  geom_point(alpha=.5) +
  geom_hline(yintercept=.05, linetype="dashed") +
  geom_vline(xintercept=.05, linetype="dashed") +
  theme_bw()

## plot accuracy vs. pvalue threshold
pr |>
  ggplot(aes(x=pvalue_threshold, y=accuracy, 
             color=as.factor(ref_qcut))) +
  geom_line() +
  geom_point(alpha=.5) +
  geom_hline(yintercept=.05, linetype="dashed") +
  geom_vline(xintercept=.05, linetype="dashed") +
  theme_bw()


```
## Within datasets: HumanCellLandscape

```{r}
RES_HCL <- RES_MAPPED[ctd=="HumanCellLandscape"] |>
  subset(stage!="HESC")|>
  data.table::dcast.data.table(
  formula = hpo_id+cell_ontology_mapped ~ stage,
  fun.aggregate = mean,
  drop = TRUE,
  value.var = c("p","q","fold_change"))
RES_HCL <- RES_HCL[complete.cases(RES_HCL)][,test_id:=.I]
### All results
message(length(unique(RES_HCL$cell_ontology_mapped))," comparable celltypes.")
message(length(unique(RES_HCL$hpo_id))," comparable phenotypes.")

### Significant results in both CTDs
RES_SIG <- RES_HCL[q_Adult<.05 & q_Fetus<.05]
message(length(unique(RES_SIG$cell_ontology_mapped))," comparable celltypes (FDR<0.05).")
message(length(unique(RES_SIG$hpo_id))," comparable phenotypes (FDR<0.05).")
```

## All results (q<1)

```{r}
RES_HCL |>
  ggstatsplot::ggscatterstats(x="p_Adult", 
                              y="p_Fetus",
                              point.args = list(alpha=.01)) +
    ggplot2::geom_density_2d_filled(alpha=.85) 
```

## Signficant results (q<0.05)

```{r}
plot_cor(RES_SIG, 
         x="fold_change_Adult",
         y="fold_change_Fetus")

```

## Significant in only Adult OR Fetus

### Adult-only
```{r}
adult_only <- RES_HCL[q_Adult<.05 & q_Fetus>=.05] 
MSTExplorer::create_dt(adult_only)
```

### Fetus-only
```{r}
fetus_only <- RES_HCL[q_Adult>=.05 & q_Fetus<.05] 
MSTExplorer::create_dt(fetus_only)
```




# Ground truth

Compare the results to some ground truth datasets from the 
Monarch Knowledge Graph.

## Manual inspection

After manual inspection it was found that 65.6% of known causal cell types (and/or their immediate precursors) were recapitulated by our MSTExplorer results. In many instances, our results further resolved the specific cell subtype(s) underlying the phenotype (e.g. "neuron" to "visceromotor neuron", or "muscle cell" to "ventricular cardiac muscle cell").

Sperm cells could not be assessed due to their absence in both transcriptomic reference datasets.

The old results (using the non-standardised version of the DescartesHuman CTD) 
had 69.5% recovery of known phenotype-celltype links, whereas the new results (DescartesHuman + HumanCellLandscape) had 65.6% recovery.
```{r}
# lm_res[,ctd:="DescartesHuman"]
# RES <- MSTExplorer::map_celltype(lm_res)
 
# RES <- res_old
kg <- data.table::fread(here::here("data/monarch_kg_cells.csv"))
kg <- kg[grepl("HP:",from)][from %in% unique(res$hpo_id)]
message(paste(
  "Remaining:",length(unique(kg$from)),
  "phenotypes across",length(unique(kg$to)),"celltypes."
))
#### Merge results with Knowledge Graph ####
kg_res <- data.table::merge.data.table(
  kg[,c("from","to","label.to")],
  res[q<0.05],
  by.x="from",
  by.y="hpo_id") |>
  data.table::setorderv(c("from","fold_change"), c(1,-1))

#### Compute average number of celltypes / phenotype ####
kg[,list(N=data.table::uniqueN(to)), by="from"]$N|>summary()
kg_res[,list(N=data.table::uniqueN(CellType)), by="from"]$N|>summary()

missing_phenos <- setdiff(kg$from, kg_res$from)
message((1-(length(missing_phenos) / length(unique(kg$from))))*100,
        "% phenotypes recovered.")

kg_missing <- kg[from%in% missing_phenos]
res_missing <- res[hpo_id%in% missing_phenos]

#### Compute increased knowledge ####
n_before <- kg$from|>unique()|>length()
n_after <- res[q<0.05]$hpo_id|>unique()|>length()
n_before <- nrow(kg)
n_after <- nrow(res[q<0.05])
n_after/n_before
```


### Traverse HP ontology

```R
hpo <- ontologyIndex::get_OBO("https://github.com/obophenotype/human-phenotype-ontology/releases/download/v2023-10-09/hp-base.obo")
hp <- KGExplorer::get_ontology("hp")
g <- KGExplorer::ontology_to(hp,"igraph")
terms <- c("HP:0000001", "HP:0000002","HP:0002493", "HP:0002450",
           "HP:0001257","HP:0012210")
simona::shortest_distances_directed(hp, terms=terms)
simona::shortest_distances_via_NCA(hp, terms=terms)
simona::term_sim(hp, terms=terms)
igraph::are.connected(g, v1="HP:0002493", v2="HP:0002450")


make_pseudo_seurat <- function(dt,
                               col,
                               new_col=col,
                               unique_only = TRUE){
  vals <- dt[[col]]
  if(unique_only){
    vals <- unique(vals) 
  } 
  X <- Matrix::Matrix(ncol = length(vals))
  colnames(X) <- vals
  rownames(X) <- paste0("gene-",seq_len(nrow(X)))
  SeuratObject::CreateSeuratObject(
    count=SeuratObject::CreateAssayObject(X), 
    meta.data = data.frame(vals, row.names = vals) |> `colnames<-`(new_col)
    )
}
obj_list <- list(
  KG=make_pseudo_seurat(dt = kg, col = "from", new_col = "hpo_id"),
  MSTExplorer=make_pseudo_seurat(dt = RES[q<0.05], col = "hpo_id")
)  
obj_list_matched <- scOntoMatch::ontoMultiMatch(
  obj_list = obj_list,
  ont = hpo,
  anno_col = NULL,
  onto_id_col = "hpo_id")

int1 <- report_matches(obj_list = obj_list_matched, col = "hpo_id")
int2 <- report_matches(obj_list = obj_list_matched, col = "cell_ontology_mapped")
setdiff(int1, int2)
```


### Investigate missing phenotype-celltype associations

Let's dig into why these very obvious associations are not enriched in our results.
```{r}
p2g <- HPOExplorer::load_phenotype_to_genes()
p2g <- p2g[hpo_id%in% missing_phenos]
p2g[,n_genes:=length(unique(gene_symbol)),by=hpo_id]
summary(p2g$n_genes)

ctd <- MSTExplorer::load_example_ctd("ctd_DescartesHuman.rds")
ctd_genes <- rownames(ctd$level_2$specificity_quantiles)
p2g <- p2g[gene_symbol %in% ctd_genes,]
p2g[,n_genes_filtered:=length(unique(gene_symbol)),by=hpo_id]
summary(p2g$n_genes_filtered)
message(length(p2g$gene_symbol[!p2g$gene_symbol %in% ctd_genes]),
        " HPO genes missing from CTD.")
```



## Traverse cell ontologies

```r
map_ontology_ids <- function(ont,
                             from_ids,
                             to_ids,
                             method=c( "getOntoMapping",
                                       "getOntoMultiMapping"
                                      )
                             ){
  from_name <- to_name <- NULL;
  
  method <- method[1]
  from_ids <- as.character(unique(from_ids))
  to_ids <- as.character(unique(to_ids))
  if(method=="getOntoMultiMapping"){
    mapped <- scOntoMatch::getOntoMultiMapping(
      ont = ont, 
      onts = list(from_ids=from_ids,
                  to_ids=to_ids))
  } else {
     mapped <- scOntoMatch::getOntoMapping(ont = ont, 
                                           onts1 = from_ids, 
                                           onts2 = to_ids)
  }
 
  map <- data.table::data.table(from=names(mapped),
                                to=unname(mapped)) 
  map[,from_name:=scOntoMatch::getOntologyName(ont, from)[from]]
  map[,to_name:=scOntoMatch::getOntologyName(ont, to)[to]]
  return(map)
} 
```

```r
compare_results <- function(obj_list,
                            RES,
                            datasets=NULL){
  
  kg <- data.table::fread("~/Downloads/monarch_kg_cells.csv")
  obj <- SeuratObject::pbmc_small[,seq(nrow(kg))]
  obj@meta.data <- kg[,cell_type:=label.to][,cell_type_ontology_term_id:=to]
  obj_list[["Monarch_KG"]] <- obj
  
  #### Subset datasets ####
  if(!is.null(datasets)){
    obj_list <- obj_list[datasets]
    RES <- RES[ctd %in% datasets]
  }
  #### Map ####
  obj_list_matched <- scOntoMatch::ontoMultiMatch(obj_list = obj_list,
                                                 ont = ont,
                                                 anno_col = 'cell_type',
                                                 onto_id_col = "cell_type_ontology_term_id")
  report_matches(obj_list_matched, col="cell_ontology_mapped")
  
  map <- lapply(obj_list_matched, function(x)x@meta.data) |> 
    data.table::rbindlist(idcol = "dataset", fill=TRUE) |>
    dplyr::select(cell_type,cell_type_ontology_term_id,cell_ontology_mapped) |>
    unique() 
  kg_mapped <- (
    data.table::merge.data.table(
      kg,
      map,
      by.x="to",
      by.y="cell_type_ontology_term_id")
  )[,cell_type_ontology_term_id:=to]
  kg_mapped[,pheno_celltype:=paste(from,cell_ontology_mapped,sep="_")]
  RES_MAPPED <- data.table::merge.data.table(
      RES,
      map,
      by.x="cell_type_ontology_term_id",
      by.y="cell_type_ontology_term_id")
  RES_MAPPED[,pheno_celltype:=paste(hpo_id,cell_ontology_mapped,sep="_")]
  pc_intersect <- intersect(unique(kg_mapped[cell_ontology_mapped!="cell"]$pheno_celltype), 
                            unique(RES_MAPPED$pheno_celltype))
  kg_mapped <- kg_mapped[pheno_celltype %in% pc_intersect]
  RES_MAPPED <- RES_MAPPED[pheno_celltype %in% pc_intersect]
  #### What proportion of the Monarch KG is covered by the MSTExplorer results? ####
  RES_MAPPED[pheno_celltype %in% kg_mapped$pheno_celltype][,list(
    TP=sum(unique(pheno_celltype) %in% unique(kg_mapped$pheno_celltype)),
    n_res_kg=length(unique(kg_mapped$pheno_celltype)),
    n_res=length(unique(pheno_celltype)),
    n_res_sig=length(unique(pheno_celltype[q<.05])),
    n_res_sig_p=length(unique(pheno_celltype[p<.05]))
    ), by=c("ctd","stage")]
}



reports <- lapply(stats::setNames(ctd_names,ctd_names), function(x){
  compare_results(obj_list,
                  RES,
                  datasets=x)
})

compare_results(obj_list,
                RES)

```
# Session info

<details>

```{r}
utils::sessionInfo()
```

</details>
<hr>
